---
title: "Penguins"
author: "Azriella Turkdogan"
date: "2024-12-11"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Penguins

This R Markdown document is made using the "palmerpenguins" package as a data set to show data visualization, and hypothesis testing, and data analysis. Below is the output for loading the packages used (tidyverse, lme4, palmerpenguins, car, multcomp).

```{r data}
library(tidyverse)
library(lme4)
library(palmerpenguins)
library(car)
library(multcomp)

data(package = 'palmerpenguins')
head(penguins)
```

## Data Exploration

### Boxplots

I used boxplots as a way to visualize what is happening with the variables. I included species, sex, body mass, and flipper length for variables.

```{r boxplots - body mass}
ggplot(data=na.omit(penguins), aes(x=species, y= body_mass_g, fill=sex))+
  geom_boxplot()+
  xlab("Species")+
  ylab("Body Mass (grams)")+
  ggtitle("Penguins Species vs. Body Mass")
```

There seems to be differences shown on body mass vs. the penguin species in the boxplot above. Gentoo especially seems to have a higher mass compared to the other two species. An anova test will be used to test the hypothesis of if the average mass of each species is significantly different. The null hypothesis will be that the species do not have significantly different average body masses.

```{r boxplot - flipper length}
ggplot(data=na.omit(penguins), aes(x=species, y= flipper_length_mm, fill=sex))+
  geom_boxplot()+
  xlab("Species")+
  ylab("Flipper Length (mm)")+
  ggtitle("Penguins Species vs. Flipper Length")
```

The flipper length between penguin species appears to be differentin the boxplot above. Gentoo seems to have the highest average flipper length among the species, and Adelie appears to have the lowest average flipper length. An anova test will be used to test the hypothesis for if the average flipper lengths of the penguin species are significantly different. The null hypothesis will be that the species do not have significantly different mean flipper lengths.

### Scatter Plot

I used a scatter plot to look at how body mass and flipper length change based on the species. I also included sex for visualization purposes.

```{r scatterplot}
ggplot(data=na.omit(penguins), aes(x=body_mass_g, y= flipper_length_mm, color=species))+
  geom_point(aes(shape=sex))+
  geom_smooth(data = penguins, 
              method ="lm", alpha = 0, show.legend = FALSE, linetype = 'solid') +
  xlab("Penguin Mass (grams)")+
  ylab("Flipper Length (mm)")+
  ggtitle("Penguin Weight vs. Flipper Length")
```

The scatterplot above shows linear trends between the penguin masses and the flipper length. With this scatterplot, a hypothesis was formed that there is a significant linear relationship between body mass and flipper length. The null hypothesis is there is no linear relationship between body mass and flipper length. Linear Models will be used to test these hypotheses.

## Hypothesis Testing

### Anova Testing

An anova test and a Tukey test was used to test the hypothesis for if the means between species for flipper length and body mass are significantly different

```{r ANOVA - flipper length}
#Anova
anova_flipper = aov(penguins$flipper_length_mm ~ penguins$species, data =penguins)
summary(anova_flipper)

#Check of normality assumption
par(mfrow = c(1, 2)) # combine plots
# histogram
hist(anova_flipper$residuals, main = "Flipper Length Distribution", xlab = "flipper length residuals")
# QQ-plot
qqPlot(anova_flipper$residuals,id = FALSE, ylab = "flipper length residuals", main = "qqplot")
```

The flipper length Anova test showed a low p-value (\<2e-16) and a large F-value of 594.8. This indicates that the averages between penguin species are significantly different.

The flipper length distribution shows a normal distribution on the histogram. The qqplot is not a perfect line, but it also shows close to normal distribution.

```{r Tukey - Flipper Length}

#Tukey Test
TukeyHSD(anova_flipper, conf.level=.95) 
plot(TukeyHSD(anova_flipper, conf.level=.95), las = 2)
```

From a Tukey test with a 95% confidence level, it can be seen that the adjusted p-values are all zero. This supports the hypothesis that all of the penguin species have significantly different flipper length averages from each other. The null hypothesis that the flipper lengths are on average the same between species is rejected.

```{r ANOVA - body mass}
#Anova
anova_mass = aov(penguins$body_mass_g ~ penguins$species, data =penguins)
summary(anova_mass)

#Check of normality assumption
par(mfrow = c(1, 2)) # combine plots
# histogram
hist(anova_mass$residuals, main = "Body Mass Distribution", xlab = "body mass residuals")
# QQ-plot
qqPlot(anova_mass$residuals,id = FALSE, ylab = "body mass residuals", main = "qqplot")
```

The body mass Anova test showed a low p-value (\<2e-16) and a large F-value of 343.6. This indicates that the averages between penguin species are significantly different.

The body mass distribution shows a normal distribution on the histogram, and the qqplot also shows normal distribution.

```{r Tukey - body mass}
#Tukey Test
TukeyHSD(anova_mass, conf.level=.95) 
plot(TukeyHSD(anova_mass, conf.level=.95), las = 2)
```

A Tukey test of body mass with 95% confidence level shows that the species Chinstrap and Adelie do not have significantly different means (p=0.88). However, Gentoo and Chinstrap, as well as Gentoo and Adelie have significantly different mean body masses (p=0). The relationship between chinstrap and Adelie fails to reject the null hypothesis that the penguin body masses are the same on average.

### Linear Models

Linear models were used to test for linearity between penguins' body mass and flipper length.

```{r Adelie - Linear Model}
adelie <- filter(penguins, penguins$species == "Adelie")
adelie_lm <- lm(adelie$flipper_length_mm ~ adelie$body_mass_g)
coeff=coefficients(adelie_lm)
eq1 = paste0("y = ", round(coeff[2],3), "x + ", round(coeff[1],3), " mm/g") #linear equation

###Plot Linear Model
plot(adelie$body_mass_g, adelie$flipper_length_mm, xlab = "Body Mass (g)",
     ylab = "Flipper Length (mm)", main = "Adelie: Flipper Length vs. Weight")
abline(adelie_lm)
mtext(eq1, 3, line=0)
summary(adelie_lm)

```

The linear model for the Adelie penguins shows a significant linear relationship with a slope of 0.00667 mm/gram (p = 1.34e-09). This model has 149 degrees of freedom and a low R-squared value at 0.214, indicating that body mass has a weak correlation with flipper length.

```{r Chinstrap - Linear Model}
chinstrap <- filter(penguins, penguins$species == "Chinstrap")
chinstrap_lm <- lm(chinstrap$flipper_length_mm ~ chinstrap$body_mass_g)
coeff2=coefficients(chinstrap_lm)
eq2 = paste0("y = ", round(coeff2[2],3), "x + ", round(coeff2[1],3), " mm/g") #linear equation

###Plot Linear Model
plot(chinstrap$body_mass_g, chinstrap$flipper_length_mm, xlab = "Body Mass (g)",
     ylab = "Flipper Length (mm)", main = "Chinstrap: Flipper Length vs. Weight")
abline(chinstrap_lm)
mtext(eq2, 3, line=0)
summary(chinstrap_lm)
```

The linear model for the Chinstrap penguins shows a significant linear relationship with a slope of 0.0119 mm/gram (p = 3.75e-09). This model has 66 degrees of freedom and a moderate R-squared value at 0.4027, indicating that body mass has a moderate correlation with flipper length.

```{r Gentoo Linear Model}
gentoo <- filter(penguins, penguins$species == "Gentoo")
gentoo_lm <- lm(gentoo$flipper_length_mm ~ gentoo$body_mass_g)
coeff3=coefficients(gentoo_lm)
eq3 = paste0("y = ", round(coeff3[2],3), "x + ", round(coeff3[1],3), " mm/g") #linear equation

###Plot Linear Model
plot(gentoo$body_mass_g, gentoo$flipper_length_mm, xlab = "Body Mass (g)",
     ylab = "Flipper Length (mm)", main = "Gentoo: Flipper Length vs. Weight")
abline(gentoo_lm)
mtext(eq3, 3, line=0)
summary(gentoo_lm)
```

The linear model for the Gentoo penguins shows a significant linear relationship with a slope of 0.00904 mm/gram (p = 2.2e-16). This model has 121 degrees of freedom and a high R-squared value at 0.4937, indicating that body mass has a high correlation with flipper length.

The hypothesis that there is a linear relationship between penguin flipper length and body mass is supported by the linear models above for all three species of penguins. The null hypothesis that there is no linear relationship between body mass and flipper length is rejected.

#### Citations

```         
R Core Team (2024). _R: A Language and Environment for Statistical Computing_. R   Foundation for Statistical Computing, Vienna, Austria.   https://www.R-project.org/.
```

```         
 Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, Hayes   A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E, Bache SM, Müller K, Ooms J,   Robinson D, Seidel DP, Spinu V, Takahashi K, Vaughan D, Wilke C, Woo K, Yutani H   (2019). “Welcome to the tidyverse.” _Journal of Open Source Software_, *4*(43),   1686. doi:10.21105/joss.01686 https://doi.org/10.21105/joss.01686.
```

```         
Douglas Bates, Martin Maechler, Ben Bolker, Steve Walker (2015). Fitting Linear   Mixed-Effects Models Using lme4. Journal of Statistical Software, 67(1), 1-48.   doi:10.18637/jss.v067.i01.
```

```         
Horst AM, Hill AP, Gorman KB (2020). palmerpenguins: Palmer Archipelago   (Antarctica) penguin data. R package version 0.1.0.   https://allisonhorst.github.io/palmerpenguins/. doi: 10.5281/zenodo.3960218.
```

```         
Fox J, Weisberg S (2019). _An R Companion to Applied Regression_, Third edition.   Sage, Thousand Oaks CA. https://www.john-fox.ca/Companion/. 
```

```         
Hothorn T, Bretz F, Westfall P (2008). “Simultaneous Inference in General   Parametric Models.” _Biometrical Journal_, *50*(3), 346-363. 
```
